---
title: "2024_04_16_MAF_exploration"
format: html
editor: source
---

Now that we have the MAF file setup correctly, we can do some exploration based off of the maftools vignette

# Libraries

```{r}
library(tidyverse)
library(maftools)
# library(plyranges)
# library(vcfR)
```

# Directories

```{r}
data.dir <- "2024_04_15_results/"

results.dir <- "2024_04_16_results/"

if(!exists(results.dir)){
  dir.create(results.dir)
}
```


# Import MAF object

```{r}
ds.maf <- maftools::annovarToMaf(paste0(data.dir, "multianno/all_multianno.txt"),
                                 MAFobj = T,
                                 tsbCol = "sample")
```

# Wrangling

```{r}
head(ds.maf@data)

levels(ds.maf@data$Tumor_Sample_Barcode)
```

# MAF Summary table functions

```{r}
getSampleSummary(ds.maf)
```

```{r}
temp <- getGeneSummary(ds.maf)
```

```{r}
getClinicalData(ds.maf)
```

```{r}
getFields(ds.maf)
```

```{r}
# write.mafSummary(maf = ds.maf, basename = ds.maf)

# Error in as.character.default(new("MAF", data = list(Chromosome = c("chr1",  : 
#   no method for coercing this S4 class to a vector

?write.mafSummary
```

# Visualization

## Summary plot

```{r}
?plotmafSummary
```

```{r}
plotmafSummary(ds.maf, rmOutlier = T, 
               addStat = "median", 
               dashboard = T, 
               titvRaw = F,
               showBarcodes = T)
```

## Oncoplot

```{r}
?oncoplot
```

```{r}
oncoplot(maf = ds.maf, top = 15,
              showTumorSampleBarcodes = T,
              )
```

## Transitions and Transversions

```{r}
?plotTiTv
```

```{r}
ds.maf.titv <- titv(maf = ds.maf,
                    plot = T,
                    useSyn = T)

plotTiTv(ds.maf.titv,
         showBarcodes = T)
```

## Lollipop plots

```{r}
?lollipopPlot
```

```{r}
lollipopPlot(maf = ds.maf,
             gene = "ABCA7",
             AACol = "AAChange.refGene",
             labelPos = "all"
             )
```

```{r}
lollipopPlot(maf = ds.maf,
             gene = "AMER3",
             AACol = "AAChange.refGene",
             labelPos = "all")

lollipopPlot(maf = ds.maf,
             gene = "AMER3",
             AACol = "aaChange",
             labelPos = "all")
```

##

```{r}
plotProtein(gene = "AMER3", refSeqID = "NM_001105193")
```

```{r}
brca <- system.file("extdata", "brca.maf.gz", package = "maftools")
brca <- read.maf(maf = brca, verbose = F)
```

```{r}
subset first!!!

rainfallPlot(maf = ds.maf, detectChangePoints = T, pointSize = 0.4)
```


```{r}
ds.maf.mutload <- tcgaCompare(maf = ds.maf,
                              cohortName = "DS MAF",
                              logscale = T,
                              capture_size = 50)
```


## Plotting VAF

```{r}
plotVaf(maf = ds.maf,
        vafCol = )

# We dont have t_vaf so I'm not sure if we can use another variable for this plot
```


## Copy number analysis

We did not run the data through the GITIC program, so we don't have the necessary input files to do this.

# Somatic Interactions

```{r}
somaticInteractions(ds.maf, top = 20, pvalue = c(0.05, 0.1))
```

Detecting cancer driver genes based on positional clustering

cite OncoriveCLUST article 7 if this winds up being useful

```{r}
?oncodrive
```

```{r}
ds.maf.sig <- oncodrive(maf = ds.maf,
                        AACol = "aaChange",
                        minMut = 5,
                        pvalMethod = "zscore")

head(ds.maf.sig)
```

```{r}
?plotOncodrive
```


```{r}
plotOncodrive(res = ds.maf.sig,
              fdrCutOff = 0.1,
              useFraction = F,
              labelSize = 1)
```

After arranging the ds.maf.sig by total mutations, I see that this outlier cluster is the TAF1L gene.

```{r}
lollipopPlot(maf = ds.maf,
             gene = "TAF1L",
             AACol = "aaChange",
             labelPos = "all")
```

Strange how this was missed in the summary plot. It seems to have more mutations that the others, though only present in 4 of the 6 samples. A quick google of the gene shows it may be sex linked to males only???

Either way, the pvalues for any of the genes are not significant. We'll have to delve into this more later.

## pfam domains

```{r}
ds.maf.pfam <- pfamDomains(maf = ds.maf,
                           AACol = "aaChange",
                           top = 10
                           )
```

```{r}
temp <- ds.maf.pfam$proteinSummary[,1:7, with = F]
```


```{r}
ds.maf.pfam$domainSummary[,1:3, with = F]
```




## Drug-Gene Interactions

This functions checks for drug/gene interactions and gene druggability info compied from Drug Gene Interaction Database (dgidb.org)

```{r}
maftools::drugInteractions(maf = ds.maf,
                           fontSize = 0.75)
```

```{r}
muc17.dgi <- drugInteractions(genes = "MUC17", drugs = T)
```

Trying ABCA7, IFITM1, and MUC17 all got the result "No claimed drugs" so not sure how to use this table at the moment.

## Oncogenic signaling pathways

```{r}
pathways(maf = ds.maf,
         plotType = "treemap")
```


```{r}
pws <- pathways(maf = ds.maf,
         plotType = "bar")
```


```{r}
plotPathways(maf = ds.maf,
             pathlist = pws)
```












```{r}
sessionInfo()
```







